One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-15 paper/response.txt

I thank the reviewer for a timely report and for providing important 
comments. I have re-structured the 
paper to re-orientate this work in order to address
the issues raised by the reviewer. The revisions are made in bold,
which I further explain below in chronological order. Response to the
reviewer's specific comments follow afterwards.

@@

This paper describes a potentially destabilising effect of adopting
a locally isothermal equation of state. The fact that  
linear disturbances do not generally conserve angular momentum in
locally isothermal discs has not been appreciated. I believe it is 
worth pointing out this property of locally isothermal disks given 
that such models are still widely used to model protoplanetary discs.

The basic idea behind the background torque is now presented first. 
The numerical simulations then serve to demonstrate
this destabilisation mechanism to make one-armed spiral patterns 
grow in locally isothermal discs.

Introduction:

In accordance with the reviewer's concerns, I have removed references
to dead zones in this section. I defer this issue
to the discussions section.

I now simply give a brief overview of possible origins of spiral 
structure in circumstellar discs. I highlight previous works on m=1 or
eccentric modes, since the present simulations give one-armed
spirals. I then state that this paper presents a distinct way to grow
one-armed spirals by exploiting the non-conservation of angular
momentum of linear disturbances in locally isothermal disks.

@@

The new introduction prevents the reader from confusing the results of  
this work with proper gravitational instabilities.

Section 2: Governing equations

I have removed the description of the specific disc models adopted in
the numerical simulations from this section in order to allow a
more general discussion of linear density waves in the next section.

@@

I now describe the destabilising mechanism before presenting numerical
simulations. I also extended the description to general
non-axisymmetric disturbances.

At the end of Section 3.4, I emphasize the fact that destabilisation
by the background torque is not directly associated with self-gravity.

Section 4: Numerical simulations

I describe the specific disc models and numerical codes used to
demonstrate the background torque in action.

@@

The results on variable temperature profiles (Section 5.2) now follows
the reference case (Section 5.1).

Section 7: Discussion

Here I address most of the reviewer's comments.

I reiterate the one-armed spiral instability is not 
self-gravitational in the usual sense. 
The radially-structured disc models adopted here permits trapped 
one-armed modes because of disc self-gravity, but it does not grow by 
exerting gravitational torques on other parts of the disc (which 
underpins classic gravitational instabilities).

I checked this by performing a new simulation with a lower disc mass 
such that Q>2.5 everywhere in the disc. Then classic
non-axisymmetric gravitational instabilities should not occur. However,
this model still develops the one-armed spiral because of the forced
temperature gradient, but it grows more slowly.

@@

The calculations for this section has been moved in
Appendix C. Here, I state the finding that co-rotation and Lindblad
resonances are not important for the one-armed spiral pattern in our
disc models. This supports the notion that the instability is not
gravitational.

Section 7.3:

Here I discuss possible relevance of the results to protoplanetary
discs.

I state that the destabilising mechanism --- the background
torque --- exists in the presence of a fixed temperature
gradient. (It only vanishes identically for strictly 
adiabatic/barotropic discs, see Appendix A.) This in turn requires the
system to return to its initial temperature profile quickly when perturbed.

@@

The simulations themselves consider an isothermal or locally
isothermal equation of state only.  Gammie (2001) and other work since
have shown that the evolution of self-gravitating discs depends both
on the stability criterion and on the balance between heating and
cooling.  If the cooling rate is fast, it is likely to fragment,
otherwise it will settle into a quasi-steady, self-gravitating state.  
************************************************************************
The issue of fragmentation due to rapid cooling, as implied by assuming a 
locally isothermal equation of state, is now mentioned in Section 7.3.1.

*************************************************************************

When considering isothermal simulations, the likely evolution is one
of two things.  It's either initially so unstable that the disc
quickly fragments, or it's initially stable enough that any initial
perturbations simply drives the system towards being even more stable
and the instability quickly weakens through rapid mass transfer (see
the work by Laughlin & Bodenheimer 1994/1996).  There is no real way
to drive a long-lived quasi-steady, non-fragmenting, self-gravitating
state because this is only possible if the system can remain
quasi-stable though balancing heating and cooling (see, for example,
Lodato & Rice 2004).  
************************************************************************
Our disc models have radial structure. It is not clear how to  
apply results obtained the above studies that use smooth discs to the 
present work.

In a smooth disc, non-axisymmetric gravitational instabilities can 
operate when Q<1.5-1.7. This is satisfied in our reference disc model, 
but this region is rather narrow, about 10 scale heights wide within the 
surface density bump. It is not obvious that the high-m modes 
observed early on in this simulation is a true gravitational instability 
(Fig. 3, left panel), because the radial disc structure may permit 
other hydrodynamic shear instabilities (Lovelace et al. 1999).

In any case, the high-m modes saturate at small amplitude (<0.01) and no 
fragmentation is seen (Fig. 2), even though the discs are (locally)
isothermal and min(Q) ~ 1.3. Thus, our models concern 
self-gravitating, but (effectively) gravitationally stable discs
(Stamatellos &  Whitworth 2008). Eventually a one-armed spiral grows,
but as explained above, this is not a gravitational instability in the
usual sense.

Durisen, Hartquist & Pickett (2008) and Rogers & Wadsley (2012) 
find that fragmentation occurs at the corotation radius. However, the
one-armed spirals in our simulations have low-frequency, so its 
corotation radius is not within the spiral arm itself. This makes
fragmentation unlikely (the one-armed spiral does not grow because 
of self-gravity anyway).

@@

In fact, in the conclusions it is suggested that the conditions used
in the simulation would require a cooling < 0.1 \Omega^{-1} which,
based on the work of Gammie 2001, would quickly lead to fragmentation.
The reason it doesn't in the work considered here is that the
isothermal assumption assumes that there is some equivalently fast
heating mechanism but, in the work here, this is not provided by the
instability itself.  If it were, the simulations would become very
quickly unstable.  Also, if one looks at the work of Clarke (2009),
Rafikov (2005), and Rice & Armitage (2009) the only region of a
realistic disc where such rapid cooling is possible is in the outer
disc, but such regions are either very gravitationally unstable and
fragment, or quickly stablised by an external heating source.  I guess
it is possible that there could be a scenario where this exactly
produces the conditions assumed here, but that would seem rather
fine-tuned and it would be more likely that an external heating source
would simply stabilise the disc (unless the disc were extremely
massive and could remain unstable even at reasonably high disc
temperatures).  
*************************************************************************
The value 0.1/Omega is based on a limited number of preliminary
simulations that are currently underway. The critical thermal
relaxation time may depend on the disc parameters. I mention this in
Section 8.1.

I thank the reviewer for pointing out external heating. In Section
7.3.1 I suggest that our disc models may be appropriate to describe
the outer parts of protoplanetary discs that are subject to external
irradiation such that they are gravitationally stable.

Note that stabilisation by external irradiation against  
self-gravitational instabilities does not necessarily imply
stablisation against the one-armed spiral instability described in
this paper.

**************************************************************************

When considering the reference run discussed on page 5, my suspicion
is that the behaviour can largely be explained as being a consequence
of the chosen numerical setup.  For example, the initial Q profile has
a minimum of about Q = 2, so the disc is unlikely to become very
unstable.  The simulation is seeded with an initial high-m
perturbation that decays quite quickly and is then replaced by a
long-lived m=1 spiral mode that continues to grow and in which the
growth rate depends on the imposed temperature profile.  
***************************************************************************
The minimum Q in the reference run is between 1.3 and 1.4 which,
according to previous studies, should result in  
strong non-axisymmetric gravitational instabilities (Durisen et
al. 2007, Helled et al. 2013). This does not occur here, despite the
simulations being (effectively) isothermal, which should favour
GI.

@@

Fig. 2 shows that the high-m modes do not decay over the simulation
timescale, but it saturates at a small amplitude. Thus, if these
high-m modes are GI, then GI is weak in our models despite having low
Q.

One possible origin of discrepancy is the fact that our discs are
radially structured, while the aforementioned studies employ global
smooth discs, or local models which excludes any effects of
global radial disc structure. It is not obvious that results from
smooth-disc studies can be directly applied to structured discs.

It is not unreasonable to study radially-structured discs. Realistic
discs may not be described by smooth power-laws, e.g. a disc 
with an gap induced by a planet (Section 7.3.2) and transition discs.

The numerical setup was chosen to effectively demonstrate the
destabilising effect of the fixed temperature gradient (Section 5.2 and
5.3). However, its occurrence is not limited to the particular setup
presented in Section 5 and 6.

@@

I have performed additional simulations with 
higher Q and simulations with only an inner bump edge (by setting a
large R2, see Eq. 33). These simulations still developed the one-armed
spiral, but require longer runs so high-resolution/3D is not
feasible. A full parameter survey is beyond the scope of this work,
but is the subject of a follow-up study.

***************************************************************************
It's my view that if you used a realistic equation of state that
included heating and cooling, the high-m mode instability would either
persist (as suggested by Lodato & Rice 2004) or the system would
simply heat up until it became stable against the growth of any
perturbation.  The disc mass is also quite low (0.086M*) and if one
were to estimate the disc temperature using some kind of realistic
assumption about the central star mass and disc outer radius, my
suspicion is that to have the Q profile used, the outer radius would
either have to be very small (a very compact disc) or the disc
temperature would be unrealistically low.  
***************************************************************************
The high-m perturbations do in fact persist in the reference run 
(Fig. 2), but it is overshadowed by the one-armed instability.

Lodato & Rice (2004) cool their discs to zero temperature
which is balanced against shock-heating (as in many similar
studies). The systems we consider are closer to irradiated discs (Rice
et al. 2011) in which the temperature is set externally 
(e.g. Stamatellos &  Whitworth 2008, Boley 2009). I note that 
such global simulations including heating (irradiation) and cooling 
may not have been run long enough to see low-frequency modes 
which grow slowly, and thus appear stable.

The properties of the disc as a whole, i.e. its total mass and outer
radius, is not expected to be important as far as destabilisation by
the temperature gradient is concerned. These parameters do not appear in
our description of this effect (Section 3). In Section 7.3.2 I suggest
disc-planet interaction as one possible origin for the required
structure in the Q-profile to trap m=1 modes.

@@

***************************************************************************
I have a further issue with the m=1 mode.  In these simulations the
central star is fixed.  As I understand it, the m=1 mode is not
axi-symmetric and hence it would seem to me that if the central star
is fixed, the presence of a dominant m=1 mode means that the
centre-of-mass is moving in the simulations.  In fact, it seems to me
that it is possible that the fixed central star is what is driving the
m=1 mode (i.e., the central star is unable to move to keep the centre
of mass fixed and this may have implications for the growth of
instabilities in the disc).  It would certainly be interesting to know
if you could get the same result as obtained here if the central star
was allowed to move.  
***************************************************************************
The usual interpretation of m=1 gravitational instabilities is
attributed to the motion of the central star (Adams, Ruden & Shu 1989,
Shu et al. 1990, Heemskerk, Papaloizou, Savonije 1992). Moving the
star with the disc requires a somewhat massive disc, which is not the
case considered in this work.

I have performed simulations including the indirect potential, and
thus allowing the star to move, which gave very similar results to
those presented, i.e. the one-armed mode still develops. This is
mentioned in Section 7.1.

***************************************************************************
As I said at the beginning, I don't have any fundamental problems with
how the paper is written, the problem it is trying to address, or how
the assumptions and setup is explained.  The main issue is that I
think it is sufficiently unrealistic that adding any kind of energy
equation (or even allowing the central star to move) would change the
evolution of these discs so much that what is presented here would not
be observed in reality.  That isn't, by itself, a reason not to
publish this paper, but it would seem that it would benefit from a
some more discussion about whether or not such a process could
actually operate and where it could operate.  There is some attempt to
do so when discussing dead zones, but I doubt that the conditions
there are actually suitable as such a region would either be stable,
or cool very slowly and be susceptible to the growth of long-lived,
high-m-mode spiral structure. 
****************************************************************************
I have moved the issue of dead zones to Section 7.3.2. Dead
zones may have the appropriate radial structure to trap one-armed
modes. However, I state that, in agreement with the reviewer, dead
zones may not have the suitable thermodynamic conditions for the instability.

I suggest a planetary gap as another possibility. I  
highlight the transition disc around HD 142527 which shows a gap structure
at large radii (>100AU), for which observations estimate a low, 
but possibly stable Toomre Q value against self-gravity. Thus, the disc model 
considered in this work may not be too artificial.

### 2015-01-15 paper/discussion.tex
first revision done

  \subsubsection{Radial disc structure}
  In our simulations the 
  $m=1$ spiral is confined between two $Q$-barriers, where real solutions to the local
  dispersion relation is possible (Eq. \ref{wavenumber}). The
  existence of such a cavity is a result of the initial surface
  density bump (Eq. \ref{sig_bump}).  
  Thus, in our disc models the main role of disc structure 
  and self-gravity is to allow a local $m=1$ mode to be set up, which is then 
  destabilised by the background torque.

@@

  However, conditions in a dead zone may not be suitable for 
  sustaining a background torque because it may not cool/heat fast enough
  to maintain a fixed temperature profile. Recently, 
  \cite{bae14} presented sophisticated numerical models of
  dead zones, including a range of heating and 
  cooling processes, which show that dead zones developed large-scale
  (genuine) gravitational instabilities with multiple spiral
  arms. Although this does not prove absence of the background torque,
  it is probably insignificant compared to gravitational torques.

  Another possibility is a gap opened by an embedded planet. In that case $Q$
  rises rapidly towards the gap centre since it is a region of low
  surface density. This can satisfy Eq. \ref{qb_cond}. Then the inner
  edge of our bump function mimics the outer gap edge. The outer gap
  edge is then a potential site for the growth of a low-frequency
  one-armed spiral through the background torque. However, the
  locally isothermal requirement would limit this process to the
  outer disc irradiated by the central star, or that the temperature
  profile about the gap edge is set by the planet luminosity.

  Here, it is worth mentioning transition disc 
  around HD 142527, the outer parts of which displays an $m=1$
  asymmetry \citep{fukagawa13} and spiral arms 
  \citep{christiaens14} just outside a disc gap. These authors estimate 
  $Q\simeq$1---2 in the outer disc, implying self-gravity is
  important, but the disc may remain gravitationally-stable
  \citep{christiaens14}. This is a  
  possible situtation that our disc models represent. 
}

>>>>

  \subsubsection{Radial disc structure}
  In our simulations the 
  $m=1$ spiral is confined between two $Q$-barriers, where real solutions to the local
  dispersion relation is possible (Eq. \ref{wavenumber}). The
  existence of such a cavity results from the adopted initial surface
  density bump (Eq. \ref{sig_bump}).  
  Thus, in our disc models the main role of disc structure 
  and self-gravity is to allow a local $m=1$ mode to be set up, which is then 
  destabilised by the background torque.

@@

  However, conditions in a dead zone may not be suitable for 
  sustaining a background torque because it may not cool/heat fast enough
  to maintain a fixed temperature profile. Recently, 
  \cite{bae14} presented numerical models of
  dead zones including a range of heating and 
  cooling processes, which show that dead zones developed large-scale
  (genuine) gravitational instabilities with multiple spiral
  arms. Although this does not prove absence of the background torque,
  it is probably insignificant compared to gravitational torques.

  Another possibility is a gap opened by an embedded planet. In that case $Q$
  rises rapidly towards the gap centre since it is a region of low
  surface density. This can satisfy Eq. \ref{qb_cond}. Then the inner
  edge of our bump function mimics the outer gap edge. The outer gap
  edge is then a potential site for the growth of a low-frequency
  one-armed spiral through the background torque. However, the
  locally isothermal requirement would limit this process to the
  outer disc, or that the temperature 
  profile about the gap edge is set by the planet luminosity.

  Here, it is worth mentioning the transition disc 
  around HD 142527, the outer parts of which displays an $m=1$
  asymmetry \citep{fukagawa13} and spiral arms 
  \citep{christiaens14} just outside a disc gap. These authors estimate 
  $Q\simeq$1---2 in the outer disc, implying self-gravity is
  important, but the disc may remain gravitationally-stable
  \citep{christiaens14}. This is a  
  possible situation that our disc models represent. 
}

### 2015-01-15 paper/model.tex
first revision done

The important consequence of the background torque is the possibility
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and the background torque $\tbg$ is {\bf also} positive (negative),
then the local angular momentum density of the linear disturbance
{\bf will} further increase (decrease) {\bf with time}. This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc.

@@

\subsection{Angular momentum density of  non-axisymmetirc low-frequency modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right].\label{ang_mom_def}  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$, {\bf we find}
\begin{align}
  \jlin &\simeq \frac{m\Sigma\Omega}{2}\left[-m|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right]\notag\\
  & = -\frac{m\Sigma\Omega}{2}\left[ (m-1)|\bm{\xi}|^2 +  |\xi_R + \ii\xi_\phi|^2\right].
%  & \leq 0.
\end{align}
{\bf
Thus, non-axisymmetric ($m\geq1$) low-frequency modes generally have 
negative angular momentum. If the mode loses (positive) angular momentum 
to the background, then we can expect instability. We show 
below how this is possible through a forced temperature
gradient via the background torque. It is simplest to calculate
$T_\mathrm{BG}$ in the local approximation, which we review first.  
}

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}
  \frac{k}{\Omega}\frac{|\xi_R|^2}{\left[(m-1)|\bm{\xi}|^2 + |\xi_R +
      \ii\xi_\phi|^2 \right]} 
  = -\frac{dc_s^2}{dR}
  \frac{k}{m\Omega},
%  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
%   \xi_\phi|^2}. 
\end{align}
{\bf where the second equality uses $\xi_\phi \simeq
2\ii \xi_R/m$ in the local approximation, as shown in  Appendix
\ref{horizontal_displacements}.} 
Then for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{m\Omega} \sim q h
  \left(\frac{kH}{m}\right)\Omega,\label{theoretical_rate}
\end{align}
where we used $c_s\sim H\Omega\sim hR\Omega$. {\bf
  Eq. \ref{theoretical_rate} suggests that perturbations with small
  radial length-scales ($kH\gtrsim m$) are most favourable for
  destabilisation. Taking the local approximation is then
  appropriate.}

@@

\subsubsection{FARGO}
Our primary code is FARGO with self-gravity \citep{baruteau08}. This
is a popular, simple finite-difference code for 2D discs. `FARGO' refers
to its azimuthal transport algorithm, which removes the mean azimuthal
velocity of the disc, thereby permit larger timesteps than that would otherwise be allowed
by the usual Courant condition based on the full azimuthal
velocity \citep{masset00a,masset00b}.

@@

The normalised surface density with azimuthal wavenumber $m$ is
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
$m^\mathrm{th}$ mode can be characterized by 
assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$ as in linear
theory.

The total non-axisymmetric surface density is 
\begin{align}
  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
\end{align}

>>>>

The important consequence of the background torque is the possibility
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and $\tbg$ is {\bf also} positive (negative),
then the local angular momentum density of the linear disturbance
{\bf will} further increase (decrease) {\bf with time}. This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc.

@@

\subsection{Angular momentum density of  non-axisymmetric low-frequency modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right].\label{ang_mom_def}  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$, {\bf we find}
\begin{align}
  \jlin &\simeq \frac{m\Sigma\Omega}{2}\left[-m|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right]\notag\\
  & = -\frac{m\Sigma\Omega}{2}\left[ (m-1)|\bm{\xi}|^2 +  |\xi_R + \ii\xi_\phi|^2\right].
%  & \leq 0.
\end{align}
{\bf
Thus, non-axisymmetric ($m\geq1$) low-frequency modes generally have 
negative angular momentum. If the mode loses (positive) angular momentum 
to the background, then we can expect instability. We show 
below how this is possible through a forced temperature
gradient via the background torque. It is simplest to calculate
$T_\mathrm{BG}$ in the local approximation, which we review first.  
}

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}
  \frac{k}{\Omega}\frac{|\xi_R|^2}{\left[(m-1)|\bm{\xi}|^2 + |\xi_R +
      \ii\xi_\phi|^2 \right]} 
  = -\frac{dc_s^2}{dR}
  \frac{k}{m\Omega},
%  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
%   \xi_\phi|^2}. 
\end{align}
{\bf where the second equality uses $\xi_\phi \simeq
2\ii \xi_R/m$ for low-frequency modes in the local approximation, as shown in  Appendix
\ref{horizontal_displacements}.} 
Then for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{m\Omega} \sim q h
  \left(\frac{kH}{m}\right)\Omega,\label{theoretical_rate}
\end{align}
where we used $c_s\sim H\Omega\sim hR\Omega$. {\bf
  Eq. \ref{theoretical_rate} suggests that perturbations with small
  radial length-scales ($kH\gtrsim m$) are most favourable for
  destabilisation. Taking the local approximation is then
  appropriate.}

@@

\subsubsection{FARGO}
Our primary code is FARGO with self-gravity \citep{baruteau08}. This
is a popular, simple finite-difference code for 2D discs. `FARGO' refers
to its azimuthal transport algorithm, which removes the mean azimuthal
velocity of the disc, thereby permit larger time-steps than that would otherwise be allowed
by the usual Courant condition based on the full azimuthal
velocity \citep{masset00a,masset00b}.

@@

The normalised surface density with azimuthal wavenumber $m$ is
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
$m^\mathrm{th}$ mode can be characterized by 
assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$ as in linear
theory. The total non-axisymmetric surface density is 
\begin{align}
  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
\end{align}

### 2015-01-15 paper/paper.tex
first revision done

\section{Introduction}\label{intro}
An exciting development in the study of circumstellar 
discs is the direct observation of large-scale, non-axisymmetric
structures within them. These
include lopsided dust distributions 
\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
spiral arms 
\citep{hashimoto11,muto12,boccaletti14,grady13,christiaens14,avenhaus14}.

@@

{\bf
  In this work we identify a new mechanism that leads to the growth of
  one-armed spirals in astrophysical discs. We show
  that when the disc temperature is prescribed (called locally isothermal
  discs), the usual statement of the conservation of angular momentum
  for linear perturbations acquires a source term proportional to the
  temperature gradient. This permits angular momenum exchange between
  linear perturbations and the background disc. This `background
  torque' can destabilise low-frequency non-axisymmetric
  trailing waves when the disc temperature decreases outwards.       
}

@@

This paper is organised as follows. In \S\ref {model} we describe the system of interest
and list the governing equations. {\bf In \S\ref{wkb} we use linear theory  to show
  how a fixed temperature profile 
  can destabilise non-axisymmetric waves in discs.} \S\ref{methods} describes
the {\bf numerical setup and} hydrodynamic codes {\bf we use to 
  demonstrate the growth of one-armed spirals due to an imposed
  temperature gradient.} Our {\bf
  simulation} results are presented in  \S\ref{results2d} for two-dimensional (2D) {\bf
  discs} and in \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
discuss them in \S\ref{discussions}. We summarise in \S\ref{summary} 
with a discussion on speculations for future work.

>>>>

\section{Introduction}\label{intro}
An exciting development in the study of circumstellar 
discs is the direct observation of large-scale, non-axisymmetric
structures within them. These
include lopsided dust distributions 
\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14,plas14} and
spiral arms 
\citep{hashimoto11,muto12,boccaletti14,grady13,christiaens14,avenhaus14}.

@@

{\bf
  In this work we identify a new mechanism that leads to the growth of
  one-armed spirals in astrophysical discs. We show
  that when the disc temperature is prescribed (called locally isothermal
  discs), the usual statement of the conservation of angular momentum
  for linear perturbations acquires a source term proportional to the
  temperature gradient. This permits angular momentum exchange between
  linear perturbations and the background disc. This `background
  torque' can destabilise low-frequency non-axisymmetric
  trailing waves when the disc temperature decreases outwards.       
}

@@

This paper is organised as follows. In \S\ref {model} we describe the system of interest
and list the governing equations. {\bf In \S\ref{wkb} we use linear theory  to show
  how a fixed temperature profile 
  can destabilise non-axisymmetric waves in discs.} \S\ref{methods} describes
the {\bf numerical setup and} hydrodynamic codes {\bf we use to 
  demonstrate the growth of one-armed spirals due to an imposed
  temperature gradient.} Our {\bf
  simulation} results are presented in  \S\ref{results2d} for two-dimensional (2D) {\bf
  discs} and in \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
discuss them in \S\ref{discussions}. We summarise in \S\ref{summary} 
with some speculations for future work.
